---
title: "Data Sources and Cleaning"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
---

```{r setup, include=FALSE}
library(tidyverse)
library(haven)

knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  fig.align = 'center')


theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

## Data Sources
The datasets employed in our project originate from two distinct primary sources. 

For part 1, we utilized [PLACES: Local Data for Better Health](https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-County-Data-20/swc5-untb/about_data) to acquire model-based county estimates of the entire United States during the time period from 2020 to 2021. The data were provided by the CDC, Division of Population Health, Epidemiology and Surveillance Branch.The estimates encompass 36 measures, spanning health outcomes, preventive services use, chronic disease-related health risk behaviors, disabilities, and health status. We are using several health-affecting behavior variables and the prevalence of High Cholesterol to evaluate and sorting out the most correlated risk factors to the outcome at the national level.

For part 2, the resource is from the program [The National Health and Nutrition Examination Survey (NHANES)](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Cycle=2017-2020) conducted by the (CDC) for the time period of 2017 to March 2020 Pre-pandemic. It combines interviews and physical examinations to collect a wide range of health-related information. NHANES involves sampling thousand of individuals across different age groups. Unlike the county estimates from part 1, this part provides individual data measurements of demographic characteristics, our selected risk factors from part 1 and our health outcome (High Cholesterol) to generate the regression model. This program divides different types of measuremens into sub-dateset:

* **[TCHOL](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_TCHOL.htm)**: It include the measurement of serum total cholesterol levels to assess the risk of high cholesterol.
* **[P_DEMO](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_DEMO.htm)**: It provides first part of the individual demographic variables including age, gender, race, marital status, and education level.
* **[P_BMX](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_BMX.htm)**: Provides second part of the demographic variable, body mass index.
* **[P_INQ](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_BMX.htm)**: It has third part of the demographic variable, family monthly poverty level index based on participants' monthly income.
* **[P_PAQ](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_PAQ.htm#PAQ665)**: It provides the first explanatory variable, respondent-level interview data on physical activities.
* **[P_SLQ](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_SLQ.htm)**: It provides the second explanatory variable, sleep hours in weekdays.
* **[P_ALQ](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_ALQ.htm#ALQ121)**: Provides the third explanatory variable, subjects' drinking patterns in the past 12 months.
* **[P_COT](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/P_COT.htm)**: Provides the fourth explanatory variable, serum cotinine level, as the primary metabolites of nicotine to estimate the extent of tobacco use.

## Data Import and Tidy
### Part 1: PLACES: Local Data for Better Health

### Part 2: NHANES
Loading the `TCHOL` dataset with the data of response variable `total cholesterol`
```{r}
tchol_df = 
  data <- read_xpt("data/xpt/P_TCHOL.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, total_cholesterol = lbxtc) |> 
  drop_na()
```

### Loading the first demographic dataset `P_DEMO`.
```{r}
demo1_df = 
  data <- read_xpt("data/xpt/P_DEMO.XPT") |> 
  janitor::clean_names() |> 
  select(id = seqn, age = ridageyr, gender = riagendr, race = ridreth3, 
         marital_status = dmdmartz, education_level_20 = dmdeduc2) |> #only include the variables of id, age, gender, race, marital status, and education level.
  filter(age >= 18) |> # Only include the participants with age equal and greater than 18.
  mutate(
    gender = case_when(
      gender == 1 ~ "Male",
      gender == 2 ~ "Female",
      TRUE ~ as.character(gender)), #categorize variable gender.
    race = case_when(
      race %in% c(1, 2) ~ "Hispanic",
      race == 3 ~ "Non-Hispanic White",
      race == 4 ~ "Non-Hispanic Black",
      race == 6 ~ "Non-Hispanic Asian",
      race == 7 ~ "Other Race",
      TRUE ~ as.character(race)), #categorize variable race.
    marital_status = case_when(
      marital_status %in% c(77, 99, ".") ~ NA_character_,
      marital_status == 1 ~ "Married/Living with Partner",
      marital_status == 2 ~ "Widowed/Divorced/Separated",
      marital_status == 3 ~ "Never married",
      TRUE ~ as.character(marital_status)), #categorize variable marital status.
    education_level_20 = case_when(
      education_level_20 %in% c(7, 9, ".") ~ NA_character_,
      education_level_20 == 1 ~ "Less than 9th grade",
      education_level_20 == 2 ~ "9-11th grade",
      education_level_20 == 3 ~ "High school graduate/GED or equivalent",
      education_level_20 == 4 ~ "Some college or AA degree",
      education_level_20 == 5 ~ "College graduate or above",
      TRUE ~ as.character(education_level_20)) #categorize education level status.
   )

```

### Loading the second demographic dataset `P_BMX` with data of body mass index.
```{r}
bmi_df = 
  data <- read_xpt("data/xpt/P_BMX.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, bmi = bmxbmi) |> 
  drop_na()
```

### Loading the third demographic dataset `P_INQ` with data of family monthly poverty level.
```{r}
income_df = 
  data <- read_xpt("data/xpt/P_INQ.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, poverty_level = indfmmpi) |> 
  drop_na() |> 
  mutate(
    poverty_level = case_when(
      poverty_level < 1.3 ~ "Below 130% of Poverty Guidelines",
      between(poverty_level, 1.3, 1.85) ~ "Between 130% and 185% of Poverty Guidelines",
      poverty_level > 1.85 ~ "Above 185% of Poverty Guidelines",
      TRUE ~ as.character(poverty_level))
        )
```

### Left join the income_df and bmi_df on demo1_df as the dataset with all the demographic data.
```{r}
demo_df <- demo1_df |> 
  left_join(income_df, by = "id") |> 
  left_join(bmi_df, by = "id") |> 
  drop_na(age, gender, race, bmi) 
```

### Loading the first explanatory variable `physical actively` from dataset `P_PAQ`.
```{r}
activity_df = 
  data <- read_xpt("data/xpt/P_PAQ.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, paq605, paq650, paq620, paq635, paq665) |> 
  drop_na() |> 
  mutate(
    physical_activity = case_when(
      paq605 == 1 | paq650 == 1 ~ "Vigorous activity",
      paq620 == 1 | paq635 == 1 | paq665 == 1 ~ "Moderate activity",
      paq665 == 2 ~ "Light/Unknown activity",
      TRUE ~ NA_character_ )
        ) |> 
  select(id, physical_activity)
```

### Loading the second explanatory variable `sleep hours` from dataset `P_SLQ`.
```{r}
sleep_df = 
  data <- read_xpt("data/xpt/P_SLQ.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, sleep_hour = sld012) |> 
  drop_na()
```

### Loading the third explanatory variable `alcohol use` from dataset `P_ALQ`.
```{r}
alcohol_df = 
  data <- read_xpt("data/xpt/P_ALQ.XPT") |> 
  janitor::clean_names() |>
  mutate(
    alcohol_use_cat = case_when(
      alq121 %in% c(0, 10, 9, 8) ~ "Light Drinker",
      alq121 %in% c(7, 6, 5, 4) ~ "Moderate Drinker",
      alq121 %in% c(3, 2, 1) ~ "Heavy Drinker",
      alq121 %in% c(77, 99, ".") ~ NA_character_),
     alcohol_use_order = case_when(
      alq121 == 0 ~ 1,
      alq121 == 10 ~ 2,
      alq121 == 9 ~ 3,
      alq121 == 8 ~ 4,
      alq121 == 7 ~ 5,
      alq121 == 6 ~ 6,
      alq121 == 5 ~ 7,
      alq121 == 4 ~ 8,
      alq121 == 3 ~ 9,
      alq121 == 2 ~ 10,
      alq121 == 1 ~ 11)
        ) |> 
   select(id = seqn, alcohol_use_cat, alcohol_use_order) |> 
   drop_na()
```

### Loading the fourth explanatory variable `smoking` from dataset `P_COT`.
```{r}
smoking_df = 
  data <- read_xpt("data/xpt/P_COT.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, cotinine = lbxcot) |> 
  drop_na()
```

```{r}
combine_df = 
  tchol_df |> 
  left_join(demo_df, by = "id") |>
  left_join(activity_df, by = "id") |>
  left_join(sleep_df, by = "id") |>
  left_join(alcohol_df, by = "id") |>
  left_join(smoking_df, by = "id") |> 
  drop_na(age, gender, race)

write.csv(combine_df, file = "data/combine.csv")
```

